Emergence of the Southeast Asian islands as a driver for Neogene cooling

### 2020-08-13 Manuscript/Manuscript.tex
revisions checkpoint

To determine the position of paleoshorelines in the SEAI over the past 15~m.y., we use terrestrial and marine sedimentary deposits (Fig. \ref{fig:shoreline_growth}; \SI). The paleoshoreline data indicate that the Sunda-Banda Arc and New Guinea are primarily responsible for the increase in area since 15~Ma. Exhumation of the modern Sunda-Banda Arc is the result of ongoing arc-continent collision with the Australian Plate \cite{Harris2006a}. Most of Sumatra and Java along with the non-volcanic islands of the Outer Banda Arc were elevated above sea level after 5~Ma \cite{Hall2013b}. In New Guinea, emergence in the mid-Miocene is associated with collision between the Melanesian Arc and Australia's distal margin \cite{Cloos2005a}, which drove exhumation of the Irian-Marum-April Ophiolite Belt. Exhumation accelerated over the past 4~m.y. in the New Guinea Central Range due to slab-breakoff and buoyant uplift, and in eastern New Guinea due to jamming of the north-dipping subduction zone \cite{Cloos2005a}. We also include changes in areas of presently submerged continental shelves such as the Sunda Shelf that were previously exposed (\SI; Fig. S6). These tectonic drivers and others throughout the region led to progressive emergence over the past 15~m.y. that accelerated following 5~Ma (Fig. \ref{fig:shoreline_growth}B). This trend mirrors broad cooling over the Neogene that resulted in the initiation of Northern Hemisphere ice sheets (Fig. \ref{fig:shoreline_growth}C).

We use GEOCLIM to estimate \pCOtwo associated with the reconstructed subaerial extent of the SEAI at ca. 15, 10, and 5~Ma (``paleo-SEAI'' scenarios; Fig. \ref{fig:scenario_pCO2}). Because we use a climate model forced with modern geography, the position of the tectonic blocks remain fixed. Although there has been motion of these tectonic blocks since 15~Ma, they have remained within tropical latitudes such that this fixed scenario is a good approximation of the paleogeography (\SI; Fig. S9). We also test an end-member scenario, in which all islands associated with arc-continent collision in the region are removed (``removed SEAI'' scenario; Fig. \ref{fig:scenario_pCO2}).

@@

Proxy-based estimates of the magnitude and trajectory of \pCOtwo change from the Miocene to the Pliocene are variable between techniques and associated assumptions underlying their interpretation (\SI; Fig. S10). The \pCOtwo values from the 5~Ma ``paleo-SEAI'' scenario overlap with many proxy-based estimates \cite{Foster2017a} as well as values that emerge from approaches that assimilate climate and ice sheet model output with benthic \dO data \cite{van-de-Wal2011a, Berends2020a}. The modeled \pCOtwo values for 15~Ma resemble the higher end of proxy-based \pCOtwo estimates for the early to mid-Miocene, indicating that the increase in subaerially exposed land area and tectonic topography of the SEAI is sufficient to explain long-term cooling of Earth's climate over the Neogene. The \pCOtwo threshold for Antarctic glaciation is estimated to be $\sim$750~ppm with that for Northern Hemisphere glaciation being significantly lower at $\sim$280~ppm \cite{DeConto2008a}. These modeled values of decreasing \pCOtwo associated with emergence of the SEAI are therefore consistent with the record of Neogene climate with Miocene ice sheets on Antarctica \cite{Sugden1995a} followed by Northern Hemisphere ice sheets developing in the Pliocene \cite{Haug2005a} as \pCOtwo subsequently decreased.

@@

An important caveat for these estimates of \pCOtwo is that our modeling is determining the climatology in the GFDL CM2.0 model at which steady-state is achieved -- a climatology that has an associated \pCOtwo value in the model. However, climate models are variable in their response to changes in \pCOtwo. One way to summarize this variability is through the equilibrium climate sensitivity value -- the steady-state change in global mean surface air temperature associated with a doubling of \pCOtwo. A range of 1.5 to 4.5\degreesC per \pCOtwo doubling was proposed in the landmark Charney report \cite{Charney1979a} and this range was considered to be the credible interval (\textgreater66\% likelihood) in the last IPCC report \cite{Stocker2013a}. Integrating constraints both from understanding of climate feedback processes and the climate record, a recent comprehensive review estimates the 66\% probability range of climate sensitivity to be 2.6 to 3.9\degreesC per \pCOtwo doubling with a 5 to 95\% range of 2.3 to 4.7\degreesC per \pCOtwo doubling \cite{Sherwood2020a}. The equilibrium climate sensitivity associated with the CM2.0 climate models is 2.9\degreesC per \pCOtwo doubling which falls within these ranges although these ranges remain broad. An alternative way to consider the results from our analysis would be that an estimate of 572~ppm (2\times pre-industrial \pCOtwo) for the 15~Ma paleo-SEAI scenario is implying that Earth would be $\sim$2.9\degreesC warmer. If Earth's climate sensitivity is at the higher end of the probable range and higher than in the CM2.0 model, as it is in some climate models, this same amount of Neogene cooling resulting from the emergence of the SEAI could have been driven by a smaller change in \pCOtwo.

@@

However, changes in the volcanic degassing flux would have modulated changes in \pCOtwo associated with changes in global weatherability. For example, some proxy-based approaches as well as some model-data assimilation approaches estimate that mid-Miocene \pCOtwo was lower than 568~ppm (\SI; Fig. S10). Take a scenario in which \pCOtwo was 400~ppm at 15~Ma. If we assume that the \pCOtwo decrease to the pre-industrial value of 286~ppm was driven by the increase in global weatherability associated with emergence of the SEAI in conjunction with an increase in volcanic degassing which counteracts cooling by increasing the flux of \COtwo to the atmosphere (``increase in weatherability and degassing'' scenario in Fig. \ref{fig:weatherability_curves}), the volcanic degassing flux would have had to have been $\sim$7\% smaller than the pre-industrial. More robust constraints on \pCOtwo (\SI; Fig. S10) and/or volcanic degassing rates over the past 20~m.y. are needed to constrain which of the ``increase in weatherability only'' or ``increase in weatherability and degassing'' scenarios (Fig. \ref{fig:weatherability_curves}) is more representative of the mechanisms driving Neogene cooling.

>>>>

To determine the position of paleoshorelines in the SEAI over the past 15~m.y., we use terrestrial and marine sedimentary deposits (Fig. \ref{fig:shoreline_growth}; \SI). The paleoshoreline data indicate that the Sunda-Banda Arc and New Guinea are primarily responsible for the increase in area since 15~Ma. Exhumation of the modern Sunda-Banda Arc is the result of ongoing arc-continent collision with the Australian Plate \cite{Harris2006a}. Most of Sumatra and Java along with the non-volcanic islands of the Outer Banda Arc were elevated above sea level after 5~Ma \cite{Hall2013b}. In New Guinea, emergence in the mid-Miocene is associated with collision between the Melanesian Arc and Australia's distal margin \cite{Cloos2005a}, which drove exhumation of the Irian-Marum-April Ophiolite Belt. Exhumation accelerated over the past 4~m.y. in the New Guinea Central Range due to slab-breakoff and buoyant uplift, and in eastern New Guinea due to jamming of the north-dipping subduction zone \cite{Cloos2005a}. We also include changes in areas of presently submerged continental shelves such as the Sunda Shelf that were previously exposed (\SI; Fig. S7). These tectonic drivers and others throughout the region led to progressive emergence over the past 15~m.y. that accelerated following 5~Ma (Fig. \ref{fig:shoreline_growth}B). This trend mirrors broad cooling over the Neogene that resulted in the initiation of Northern Hemisphere ice sheets (Fig. \ref{fig:shoreline_growth}C).

We use GEOCLIM to estimate \pCOtwo associated with the reconstructed subaerial extent of the SEAI at ca. 15, 10, and 5~Ma (``paleo-SEAI'' scenarios; Fig. \ref{fig:scenario_pCO2}). Because we use a climate model forced with modern geography, the position of the tectonic blocks remain fixed. Although there has been motion of these tectonic blocks since 15~Ma, they have remained within tropical latitudes such that this fixed scenario is a good approximation of the paleogeography (\SI; Fig. S10). We also test an end-member scenario, in which all islands associated with arc-continent collision in the region are removed (``removed SEAI'' scenario; Fig. \ref{fig:scenario_pCO2}).

@@

Proxy-based estimates of the magnitude and trajectory of \pCOtwo change from the Miocene to the Pliocene are variable between techniques and associated assumptions underlying their interpretation (\SI; Fig. S11). The \pCOtwo values from the 5~Ma ``paleo-SEAI'' scenario overlap with many proxy-based estimates \cite{Foster2017a} as well as values that emerge from approaches that assimilate climate and ice sheet model output with benthic \dO data \cite{van-de-Wal2011a, Berends2020a}. The modeled \pCOtwo values for 15~Ma resemble the higher end of proxy-based \pCOtwo estimates for the early to mid-Miocene, indicating that the increase in subaerially exposed land area and tectonic topography of the SEAI is sufficient to explain long-term cooling of Earth's climate over the Neogene. The \pCOtwo threshold for Antarctic glaciation is estimated to be $\sim$750~ppm with that for Northern Hemisphere glaciation being significantly lower at $\sim$280~ppm \cite{DeConto2008a}. These modeled values of decreasing \pCOtwo associated with emergence of the SEAI are therefore consistent with the record of Neogene climate with Miocene ice sheets on Antarctica \cite{Sugden1995a} followed by Northern Hemisphere ice sheets developing in the Pliocene \cite{Haug2005a} as \pCOtwo subsequently decreased.

@@

An important caveat for these estimates of \pCOtwo is that our modeling is determining the climatology in the GFDL CM2.0 model at which steady-state is achieved -- a climatology that has an associated \pCOtwo value in the model. However, climate models are variable in their response to changes in \pCOtwo. One way to summarize this variability is through the equilibrium climate sensitivity value -- the steady-state change in global mean surface air temperature associated with a doubling of \pCOtwo. A range of 1.5 to 4.5\degreesC per \pCOtwo doubling was proposed in the landmark Charney report \cite{Charney1979a} and this range was considered to be the credible interval (\textgreater66\% likelihood) in the last IPCC report \cite{Stocker2013a}. Integrating constraints both from understanding of climate feedback processes and the climate record, a recent comprehensive review estimates the 66\% probability range of climate sensitivity to be 2.6 to 3.9\degreesC per \pCOtwo doubling with a 5 to 95\% range of 2.3 to 4.7\degreesC per \pCOtwo doubling \cite{Sherwood2020a}. The equilibrium climate sensitivity associated with the CM2.0 climate models is 2.9\degreesC per \pCOtwo doubling, which falls within these ranges although these ranges remain broad. An alternative way to consider the results from our analysis would be that an estimate of 572~ppm (2$\times$ pre-industrial \pCOtwo) for the 15~Ma ``paleo-SEAI'' scenario is implying that Earth would be $\sim$2.9\degreesC warmer. If Earth's climate sensitivity is at the higher end of the probable range and higher than in the CM2.0 model, as it is in some climate models, this same amount of Neogene cooling resulting from the emergence of the SEAI could have been driven by a smaller change in \pCOtwo.

@@

However, changes in the volcanic degassing flux would have modulated changes in \pCOtwo associated with changes in global weatherability. For example, some proxy-based approaches as well as some model-data assimilation approaches estimate that mid-Miocene \pCOtwo was lower than 568~ppm (\SI; Fig. S11). Take a scenario in which \pCOtwo was 400~ppm at 15~Ma. If we assume that the \pCOtwo decrease to the pre-industrial value of 286~ppm was driven by the increase in global weatherability associated with emergence of the SEAI in conjunction with an increase in volcanic degassing which counteracts cooling by increasing the flux of \COtwo to the atmosphere (``increase in weatherability and degassing'' scenario in Fig. \ref{fig:weatherability_curves}), the volcanic degassing flux would have had to have been $\sim$7\% smaller than the pre-industrial. More robust constraints on \pCOtwo (\SI; Fig. S11) and/or volcanic degassing rates over the past 20~m.y. are needed to constrain which of the ``increase in weatherability only'' or ``increase in weatherability and degassing'' scenarios (Fig. \ref{fig:weatherability_curves}) is more representative of the mechanisms driving Neogene cooling.

### 2020-08-13 Manuscript/Supplementary_Information.tex
revisions checkpoint

A related issue is that our analysis focuses on the weathering of Ca and Mg silicates. Sources of alkalinity to the ocean include the four major cations of Ca, Mg, Na and K. A common approach for the long-term carbon cycle follows that of \citet{Berner1983a} where an emphasis is placed on Ca and Mg with the weathering of Na and K silicates considered to not be a long-term sink of \COtwo. The reasoning behind this assumption is that while weathering of these silicates can lead to transient uptake of \COtwo, this uptake is balanced by release. The ultimate mechanism through which silicate weathering sequesters carbon is the incorporation of Ca+Mg carbonates into the lithosphere. Given that Na and K are not precipitating carbonate, when simulating long term steady-state solutions, as we are here, it is valid to follow the approach of \citet{Berner1983a} and not consider the Na-K-\COtwo cycles. Complexity arises given the long residence time of Na and K in seawater and the lag between the generation of alkalinity and its consumption that could modulate the carbonate system. Dynamic modeling of the effects of the Na and K cycles on the long term carbon cycle is an area ripe for future research although uncertainty in their sinks and the timescale of the processes that control them make them difficult to parameterize. An alternative approach for estimating carbon sequestration associated with silicate weathering that sought to incorporate all four cations was taken by \citet{France-Lanord1997a} who used a formulation of $\Delta \textrm{CO}_2 = \Delta \textrm{Mg} + \Delta \textrm{Ca} + 0.15 \Delta \textrm{Na} + 0.1 \Delta \textrm{K}$. This relationship, also applied by \citet{Schopka2011a}, assumes that all Ca$^{2+}$ and Mg$^{2+}$ eventually forms carbonates and applies an estimate that 20\% of K$^{+}$ and 30\% of Na$^{+}$ exchanges for Ca and Mg and ultimately produces carbonate (and considers charge balance). Different parameterizations such as that of \citet{France-Lanord1997a} could modulate our estimates, but the importance of the emergence of the SEAI on Neogene \pCOtwo would remain unchanged. The relative importance of mafic lithologies also remains given that with significantly higher Ca+Mg concentrations and subequal Na+K concentrations to intermediate/felsic lithologies they have the highest Ca+Mg+Na+K concentrations (Fig. \ref{fig:lithology_composition}).

>>>>

A related issue is that our analysis focuses on the weathering of Ca and Mg silicates. Sources of alkalinity to the ocean include the four major cations of Ca, Mg, Na and K. A common approach for the long-term carbon cycle follows that of \citet{Berner1983a} where an emphasis is placed on Ca and Mg with the weathering of Na and K silicates considered to not be a long-term sink of \COtwo. The reasoning behind this assumption is that while weathering of these silicates can lead to transient uptake of \COtwo, this uptake is balanced by release. The ultimate mechanism through which silicate weathering sequesters carbon is the incorporation of Ca+Mg carbonates into the lithosphere. Given that Na and K are not precipitating carbonate, when simulating long term steady-state solutions, as we are here, it is valid to follow the approach of \citet{Berner1983a} and not consider the Na and K cycles. Complexity arises given the long residence time of Na and K in seawater ($\sim$45--80~m.y. for Na, and $\sim$8--10~m.y. for K; \citealp{Emerson2008a, Lecuyer2016a}) and the lag between the generation of alkalinity and its consumption that could modulate the carbonate system. Dynamic modeling of the effects of the Na and K cycles on the long term carbon cycle is an area ripe for future research although uncertainty in their sinks and the timescale of the processes that control them make them difficult to parameterize. An alternative approach for estimating carbon sequestration associated with silicate weathering that sought to incorporate all four cations was taken by \citet{France-Lanord1997a} who used a formulation of $\Delta \textrm{CO}_2 = \Delta \textrm{Mg} + \Delta \textrm{Ca} + 0.15 \Delta \textrm{Na} + 0.1 \Delta \textrm{K}$. This relationship, also applied by \citet{Schopka2011a}, assumes that all Ca$^{2+}$ and Mg$^{2+}$ eventually forms carbonates and applies an estimate that 20\% of K$^{+}$ and 30\% of Na$^{+}$ exchanges for Ca and Mg and ultimately produces carbonate (and considers charge balance). Different parameterizations such as that of \citet{France-Lanord1997a} could modulate our estimates, but the importance of the emergence of the SEAI on Neogene \pCOtwo would remain unchanged. The relative importance of mafic lithologies also remains given that with significantly higher Ca+Mg concentrations and subequal Na+K concentrations to intermediate/felsic lithologies they have the highest Ca+Mg+Na+K concentrations (Fig. \ref{fig:lithology_composition}).
